The Precambrian paleogeography of Laurentia

### 2021-03-07 Manuscript/Laurentia_Paleogeo_Manuscript.tex
more TR comment-related edits

In this chapter, I focus on the compilation of paleomagnetic poles developed through the Nordic Paleomagnetism Workshops with some additions and modifications (Fig. \ref{fig:Laurentia_poles} and Table 2). The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them grades meant to convey the confidence that the community has in these results \citep{Evans2021a}. While many factors associated with paleomagnetic poles can be assessed quantitatively through Fisher statistics and the precision of geochronological constraints, other aspects such as the degree to which available field tests constrain the magnetization to be primary require expert assessment. The categorizations used by the expert panel are `A' and `B' with the last panel meeting occurring in Fall 2017 in Leirubakki, Iceland \citep{Brown2018a}. The `A' rating refers to poles that are judged to be of such high quality that they provide essential constraints that should be satisfied in paleogeographic reconstructions. The `B' rating is associated with poles that are judged to likely provide a high-quality constraint, but have some deficiency such as remaining ambiguity in the demonstration of primary remanence or the quality/precision of available geochronologic constraints. In addition, in this chapter I refer to select additional poles that were not given an `A' or `B' classification at the Nordic Workshops as not-rated (`NR'). These additional poles are taken from the Paleomagia database \citep{Veikkolainen2014a} in conjunction with the papers in which they were reported. Many additional poles in the Paleomagia database are valuable and should not be dismissed from being considered in paleogeographic reconstructions. However, there are ambiguities associated with many of the poles not given Nordic `A' or `B' ratings in terms of how well the nature of the remanence is constrained, including its age. For example, there are many poles that have been developed from lithologies of the Trans-Hudson orogen (e.g. \citealp{Symons2005a}. However, these poles typically lack field tests, have poor control on paleohorizontal and are hypothesized to suffer from variable overprints as discussed in \cite{Raub2008a} and \cite{DAgrella-Filho2020a}. Given the preponderance of similar directions, three representative poles from the Trans-Hudson orogen are included in the Nordic Workshop compilation (Table 2) as `B' rated poles. However, they should be used with caution given that they suffer from the same deficiencies.

There are similar issues with the rich data sets associated with intrusive and metamorphic lithologies of the Grenville Province that are the available paleomagnetic constraints for Laurentia at the Mesoproterozoic-Neoproterozoic boundary. The ages of the remanence associated with these poles is complicated by the reality that the magnetization was acquired during exhumation and associated cooling within the Grenvillian orogen. Cooling ages of deeply exhumed lithologies are more difficult to robustly constrain than the ages of remanence associated with dated eruptive units or shallow-level intrusions. As a result, the vast majority of Grenville Province poles are not given an `A' or `B' rating with the exception of the `B' rated pole from the ca. 1015 Ma Haliburton intrusions (Table 2). However, while any one of these Grenville poles could be interpreted to suffer from large temporal uncertainty, the overall preponderance of poles in a similar location at the time suggests that they need to be taken seriously within paleogeographic reconstructions of Laurentia (although an alternative view of an allochthonous origin put forward by \citealp{Halls2015a} is discussed below). In this compilation, the poles of \cite{Brown2012a} from the Adirondack highlands are used wherein the magnetic mineralogy and associated relative ages of remanence are relatively well-constrained (Table 2). An additional not-rated pole included in the present compilation is a more recently published pole for the ca. 1144 Ma Ontario lamprophyre dikes \citep{Piispa2018a} that strengthens the position of Laurentia at the time and coincides with the position of the poles from the ca. 1140 Ma Abitibi dikes \citep{Ernst1993a}. This pole will likely receive an `A' rating when assessed at the next Nordic paleomagnetism workshop. Poles from the Neoproterozoic Chuar Group of southwestern Laurentia (ca. 760 Ma)  as presented in \cite{Eyster2020a}, incorporating data from \cite{Weil2004a}, are also included.

@@

Following the amalgamation of the Archean provinces in Laurentia ca. 1.8 Ga, poles from each part of Laurentia can be considered to reflect the position of the entire composite craton. It is worth considering the possibility that poles from zones of Paleoproterozoic and Mesoproterozoic accretion could be allochthonous to the craton. \cite{Halls2015b} argued that this was the case for late Mesoproterozoic and early Neoproterozoic poles from east of the Grenvillian allochthon boundary fault. However, the majority of researchers have considered these poles to post-date major differential motion and be associated with cooling during collapse of a thick orogenic plateau developed during continent-continent collision (e.g. \citealp{Brown2012a}). Poles with a B-rating are also included in the compilation that come from Greenland, Svalbard and Scotland. These terranes were once part of contiguous Laurentia, but have subsequently been separated through translation and rifting. These poles need to be rotated into the Laurentia reference frame prior to use for tectonic reconstruction, and I apply the rotations shown in Table \ref{tab:terrane_rotations}. The Euler pole and rotation is quite well-constrained for Greenland as it is associated with recent opening of Baffin Bay and the Labrador Sea (for which the rotation of \citealp{Roest1989a} is used). The reconstruction of Scotland is associated with the opening of the Atlantic (for which the rotation employed by \citealp{Torsvik2017a} is used) which is well-constrained, but has more uncertainty associated with the Euler pole than that for Greenland. The reconstruction of Svalbard is more challenging and non-unique given a multi-stage tectonic history involving both translation within the Caledonides and subsequent rifting. The preferred Euler pole parameters of \cite{Maloof2006a} are used here for this reconstruction. This Euler rotation is designed, in particular, to honor the high degree of similarity between Tonian sediments in East Greenland and those of East Svalbard \citep{Maloof2006a, Hoffman2012a} and to reconstruct East Svalbard to be aligned with these correlative sedimentary rocks.

\begin{table}[hbt]
\caption{Rotations of separated terranes}
{\scriptsize
\begin{tabular}{|l|l|l|l|p{1.1 in}|}
  \hline
& Euler pole & Euler pole & rotation & note and \\
Terrane & longitude & latitude & angle & citation \\
\hline
Greenland & -118.5 & 67.5 & -13.8 & Cenozoic separation of Greenland from Laurentia associated with opening of Baffin Bay and the Labrador Sea \citep{Roest1989a} \\
\hline
Scotland & 161.9 & 78.6 & -31.0 & Reconstructing Atlantic opening following \cite{Torsvik2017a} \\
\hline
Svalbard & 305.0 & 81.0 & -68 & Rotate Svalbard to Laurentia in fit that works well with East Greenland basin according to \cite{Maloof2006a}\\
\hline
\end{tabular}
}
\label{tab:terrane_rotations}
\end{table}

Through the Proterozoic, there are intervals where there are abundant paleomagnetic poles that constrain Laurentia's position and intervals when the record is sparse (shown colored by age in Fig. \ref{fig:Laurentia_poles}). To further visualize the temporal coverage of the poles and to summarize the motion, implied paleolatitudes for an interior point on Laurentia are shown in Figure \ref{fig:Laurentia_paleolatitude}. The ages of the utilized paleomagnetic poles are also shown in comparison to the simplified summary of tectonic events in Figure \ref{fig:tectonic_history}. Both collisional and extensional tectonism can result in the formation of lithologies that can be used to develop paleomagnetic poles either as a result of basin formation, magmatism or both. In addition, intraplate magmatism resulting from plume-related large-igneous provinces (LIPs) can lead to paleomagnetic poles in periods that are otherwise characterized by tectonic quiescence (e.g. the ca. 1267 Ma Mackenzie LIP; Fig. \ref{fig:tectonic_history}). Intracontinental rifts have led to the highest density of poles both in the case of the ca. 1.4 Ga Belt Supergroup and the ca. 1.1 Ga Midcontinent Rift (Figs. \ref{fig:Laurentia_map} and \ref{fig:tectonic_history}). The quality and resolution of the record from the Midcontinent Rift is aided by the voluminous magmatism that occurred in conjunction with basin formation that enables the development of a well-calibrated apparent polar wander path \citep{Swanson-Hysell2019a}. The late Tonian Period also has a number of poles including the Gunbarrel LIP (ca. 780 Ma) and Franklin LIP (ca. 720 Ma), as well as similarly-aged sedimentary rocks from western Laurentia basins \citep{Eyster2020a}. Overall, there is internal consistency among the paleomagnetic poles within intervals for which there is high-resolution coverage. These data result in progressive paths such as ascending to the Logan Loop \citep{Robertson1971a}, down the Keweenawan Track \citep{Swanson-Hysell2019a} to the Grenville Loop \citep{McWilliams1975a} prior to a temporal gap before the late Tonian (ca. 775 to 720 Ma) path \citep{Eyster2020a}.

@@

An exception to this overall agreement between coeval poles from Greenland and mainland Laurentia occurs ca. 1382 Ma. There are poles of this age from Greenland associated with the Zig-Zag Dal basalts and related intrusions \citep{Marcussen1983a, Abrahamsen1987a}. However, these poles are in a distinct location from poles of similar age associated with the Belt Supergroup (e.g. the McNamara Formation and Pilcher/Garnet Range and Libby Formations; \citealp{Elston2002a}). Additionally, the older Belt Supergroup poles form a more southerly population than time-equivalent poles from elsewhere in Laurentia such as the Mistastin Pluton. There are potential complications associated with the Belt Supergroup being exposed within thrust sheets with significant Mesozoic and Cenozoic deformation. However, vertical axis rotations of the Belt region are not able to bring the Belt poles into agreement with those from Laurentia or Greenland nor is translation away from the craton. Another potential complication is that the remanence used for the development of the Belt Supergroup resides in hematite. As a result, there is the potential for inclination-flatting within the sedimentary rocks from which poles are developed. However, applying a moderate inclination factor of $f=0.6$ also does not bring the poles into congruence with the Zig-Zag Dal basalts. There is the potential that the hematite could be the result of post-depositional oxidation --- the remanence of the Purcell lavas pole is also held by hematite such that it is a chemical remanent magnetization (potentially acquired soon after eruption) rather than being a thermal remanent magnetization held by magnetite \citep{Elston2002a}. However the overall coherency of the pole directions from the Belt Supergroup and the presence of reversals as interpreted from antipodal directions has been taken as evidence that the remanence is primary \citep{Elston2002a}. At present, it is unclear which poles are a better representation of Laurentia's position ca. 1400 Ma.

The interval of the record when there are the most significant inconsistencies between poles of similar age is the Ediacaran Period at the end of the Neoproterozoic Era (Figs. \ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitude}).  Between 583 and 565 Ma, paleomagnetic poles imply both low-latitude and high-latitude positions of Laurentia  (Fig. \ref{fig:Laurentia_paleolatitude}). This conflicting record is a longstanding problem and has led to the presentation of both high-latitude and low-latitude Laurentia paleogeographic reconstructions at the time (e.g. \citealp{Pisarevsky2001a,Li2008a}). One explanation for these variable pole positions is that they are the result of large-scale oscillatory true polar wander in the Ediacaran where rapid rotation of the entire silicate Earth influenced poles in Baltica and West Africa as well \citep{McCausland2007a, Robert2017a}. Paleodirectional data from single feldspar crystals from the Sept-\^Iles layered intrusion led \cite{Bono2015a} to interpret the lower inclination (and therefore lower latitude) direction from the intrusion (the one included as the ca. 565 Ma Sept-\^Iles pole in Table 2; \citealp{Tanczyk1987a}) as the primary thermal remanent magnetization.  \cite{Bono2015a} interpret steeper directions also recovered from the intrusives as the result of remagnetization. They suggest that other steep magnetizations from Ediacaran Laurentia plutonic rocks, such as that observed in the ca. 583 Ma Baie des Mountons complex (the A group of \cite{McCausland2011a} in Table 2), are also the result of remagnetization. The lower inclination Baie des Mountons complex B Group directions result in a pole that is indistinguishable from the lower inclination  Sept-\^Iles intrusives pole. Another possibility discussed in the literature is that the lack of congruency between poles in this time interval is due to a particularly weak and non-dipolar geomagnetic field \citep{Abrajevitch2010a, Halls2015a, Bono2019a}. Data from the ca. 585 Ma Grenville dyke swarm of Laurentia, that are interpreted as primary, reveal $\sim$90\textdegree\ differences in direction within dikes dated within 2.5 $\pm$ 0.9 million years of one another \citep{Halls2015a}. The rates of $>$26\textdegree /Myr ($>$288 cm/yr) implied if these data are interpreted as resulting from plate motion or true polar wander were considered as dynamically implausible by \cite{Halls2015a} leading the authors to favor a deviation from axial dipolar behavior as the explanation for disparate Ediacaran directions. Estimates of magnetic paleointensity in these Grenville dikes are anomalously weak \citep{Thallner2020a} as are data from coeval volcanics in Ukraine \citep{Shcherbakova2019a} which could support an anomalous deviation from stable axial dipolar geomagnetic field behavior at the time as interpreted by \cite{Halls2015a}.  Regardless of mechanism, the Ediacaran data stand out as anomalous relative to the coherency of the rest of the poles in the compiled record for Laurentia (Fig. \ref{fig:Laurentia_paleolatitude}).

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../Figures/Fig6_Laurentia_reconstructions.pdf}
\caption{\textbf{Paleogeographic reconstructions of Laurentia at time intervals through the Proterozoic.} These reconstructions use the simple Laurentia pole interpolation model that is shown in Figure \ref{fig:Laurentia_paleolatitude} to reconstruct the tectonic elements of \cite{Whitmeyer2007a} shown in Figure \ref{fig:Laurentia_map}. Modern coastlines are maintained in these polygons so that the rotated orientations can be interpreted by the reader in comparison to Figure \ref{fig:Laurentia_map}. Paleomagnetic poles within 25 million years of each reconstruction time are plotted. All reconstructions have poles within such a time frame that provide constraints with the exception of the 850 Ma reconstruction which is shown faintly given this relative uncertainty in Laurentia's position.}
\label{fig:Laurentia_reconstructions}
\end{figure*}

@@

There is an overall lack of models in the literature for the Proterozoic with published continuous rotation parameters that can be compared to the compilation of paleomagnetic poles presented herein. The approach in the community for many years has been to publish models as snapshots at given time intervals presented in figures without publishing continuous rotation parameters, although some studies have published the Euler rotations associated with specified times. With the further adoption of software tools such as GPlates, there has been significant progress in the publication of continuous paleogeographic models constrained by paleomagnetic poles through the Phanerozoic (540 Ma to present; e.g. \citealp{Torsvik2017a}).

An exception to the paucity of published continuous paleogeographic models for the Precambrian is the Neoproterozoic model of \cite{Merdith2017b} which is shown in comparison to the constraints for Laurentia in Figure \ref{fig:Laurentia_paleolatitude}. The extent to which the implied position of Laurentia in \cite{Merdith2017b} is consistent with the compiled paleomagnetic constraints can be visualized in Figure \ref{fig:Laurentia_paleolatitude}. As noted above, the development of such models is challenging and  researchers need to balance varying constraints. The focus here will be on the extent to which this model satisfies the available paleomagnetic poles for Laurentia. The model does not honor the Grenville loop (e.g. Laurentia going to moderately high southerly latitudes ca. 1000 Ma), which is a striking departure from the paleomagnetic record and standard paleogeographic models. Additionally, the implemented plate motion of Laurentia in the \cite{Merdith2017b} model strays from the younger poles of the Keweenawan Track and does not honor the Franklin LIP pole \citep{Denyszyn2009b} despite its `A' Nordic rating (Fig. \ref{fig:Laurentia_paleolatitude}). The Franklin pole is taken to be a key constraint at the Tonian/Cryogenian boundary that provides evidence both for the supercontinent Rodinia being equatorial and for ice sheets associated with the Sturtian glaciation having extended to equatorial latitudes \citep{Macdonald2010a}.

There are more published models that show snapshots and publish rotation parameters associated with given time intervals such as the Rodinia model of \cite{Li2008a} and the Mesoproterozoic model of \cite{Pisarevsky2014b}, without providing parameters for a continuous model. The position for Laurentia implied by the Euler poles given for the model snapshots of these studies are shown in Figure \ref{fig:Laurentia_paleolatitude} and can be compared to the compiled record. The figure also shows the continuous implied position of Laurentia from the late Mesoproterozoic into the early Paleozoic from the model of \citet{Li2013a} (while the model parameters were not published with that study they have now been made available by the authors). This paleogeographic model implements large oscillations including ones in the Ediacaran that result from an interpretation that steep inclinations are the result of rapid motion of Laurentia from low to high latitudes and back again. The rates of Laurentia's motion associated with these models are also summarized in Figure \ref{fig:Laurentia_paleolatitude}. Over much of the record, Laurentia's pole positions can be satisfied through motion of the continent at rates of $<$10 cm/yr with intervals of more rapid motion such as during the Keweenawan Track and in the Paleozoic (Fig. \ref{fig:Laurentia_paleolatitude}).

>>>>

In this chapter, I focus on the compilation of paleomagnetic poles developed through the Nordic Paleomagnetism Workshops with some additions and modifications (Fig. \ref{fig:Laurentia_poles} and Table 2). The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them grades to convey the confidence that the community has in these results \citep{Evans2021a}. While many factors associated with paleomagnetic poles can be assessed quantitatively through Fisher statistics and the precision of geochronological constraints, other aspects such as the degree to which available field tests constrain the magnetization to be primary require expert assessment. The categorizations used by the expert panel are `A' and `B' with the last panel meeting occurring in Fall 2017 in Leirubakki, Iceland \citep{Brown2018a}. The `A' rating refers to poles that are judged to be of such high quality that they provide essential constraints that should be satisfied in paleogeographic reconstructions. The `B' rating is associated with poles that are judged to likely provide a high-quality constraint, but have some deficiency such as remaining ambiguity in the demonstration of primary remanence or the quality/precision of available geochronologic constraints. In addition, in this chapter I refer to select additional poles that were not given an `A' or `B' classification at the Nordic Workshops as not-rated (`NR'). These additional poles are taken from the Paleomagia database \citep{Veikkolainen2014a} in conjunction with the papers in which they were reported. Many additional poles in the Paleomagia database are valuable and should not be dismissed from being considered in paleogeographic reconstructions. However, there are ambiguities associated with many of the poles not given Nordic `A' or `B' ratings in terms of how well the nature of the remanence is constrained, including its age. For example, there are many poles that have been developed from lithologies of the Trans-Hudson orogen (e.g. \citealp{Symons2005a}. However, these poles typically lack field tests, have poor control on paleohorizontal and are hypothesized to suffer from variable overprints as discussed in \cite{Raub2008a} and \cite{DAgrella-Filho2020a}. Given the preponderance of similar directions, three representative poles from the Trans-Hudson orogen are included in the Nordic Workshop compilation (Table 2) as `B' rated poles. However, they should be used with caution given that they suffer from the same deficiencies.

There are similar issues with the rich data sets associated with intrusive and metamorphic lithologies of the Grenville Province that are the available paleomagnetic constraints for Laurentia at the Mesoproterozoic-Neoproterozoic boundary. The ages of the remanence associated with these poles is complicated by the reality that the magnetization was acquired during exhumation and associated cooling within the Grenville orogen. Cooling ages of deeply exhumed lithologies are more difficult to robustly constrain than the ages of remanence associated with dated eruptive units or shallow-level intrusions. As a result, the vast majority of Grenville Province poles are not given an `A' or `B' rating with the exception of the `B' rated pole from the ca. 1015 Ma Haliburton intrusions (Table 2). However, while any one of these Grenville poles could be interpreted to suffer from large temporal uncertainty, the overall preponderance of poles in a similar location at the time suggests that they need to be taken seriously within paleogeographic reconstructions of Laurentia (although an alternative view of an allochthonous origin put forward by \citealp{Halls2015a} is discussed below). In this compilation, the poles of \cite{Brown2012a} from the Adirondack highlands for which the magnetic mineralogy and associated relative ages of remanence are relatively well-constrained are included (Table 2). An additional not-rated pole included in the present compilation is a more recently published pole for the ca. 1144 Ma Ontario lamprophyre dikes \citep{Piispa2018a} that strengthens the position of Laurentia at the time and coincides with the position of the poles from the ca. 1140 Ma Abitibi dikes \citep{Ernst1993a}. This pole will likely receive an `A' rating when assessed at the next Nordic paleomagnetism workshop. Poles from the Neoproterozoic Chuar Group of southwestern Laurentia (ca. 760 Ma)  as presented in \cite{Eyster2020a}, incorporating data from \cite{Weil2004a}, are also included.

@@

Following the amalgamation of the Archean provinces in Laurentia by ca. 1.8 Ga, poles from each part of Laurentia can be considered to reflect the position of the entire composite craton. It is worth considering the possibility that poles from zones of Paleoproterozoic and Mesoproterozoic accretion could be allochthonous to the craton. \cite{Halls2015b} argued that this was the case for late Mesoproterozoic and early Neoproterozoic poles from east of the Grenvillian allochthon boundary fault. However, the majority of researchers have considered these poles to post-date major differential motion and be associated with cooling during collapse of a thick orogenic plateau developed during continent-continent collision (e.g. \citealp{Brown2012a}). Poles with a B-rating are also included in the compilation that come from Greenland, Svalbard and Scotland. These terranes were once part of contiguous Laurentia, but have subsequently been separated through translation and rifting. These poles need to be rotated into the Laurentia reference frame prior to use for tectonic reconstruction, and I apply the rotations shown in Table \ref{tab:terrane_rotations}. The Euler pole and rotation is quite well-constrained for Greenland as it is associated with recent opening of Baffin Bay and the Labrador Sea (for which the rotation of \citealp{Roest1989a} is used). The reconstruction of Scotland is associated with the opening of the Atlantic (for which the rotation employed by \citealp{Torsvik2017a} is used) which is well-constrained, but has more uncertainty associated with the Euler pole than that for Greenland. The reconstruction of Svalbard is more challenging and non-unique given a multi-stage tectonic history involving both translation within the Caledonides and subsequent rifting. The preferred Euler pole parameters of \cite{Maloof2006a} are used here for this reconstruction. This Euler rotation is designed, in particular, to honor the high degree of similarity between Tonian sediments in East Greenland and those of East Svalbard \citep{Maloof2006a, Hoffman2012a} and to reconstruct East Svalbard to be aligned with these correlative sedimentary rocks.

\begin{table}[hbt]
\caption{Rotations of separated terranes}
{\scriptsize
\begin{tabular}{|l|l|l|l|p{1.1 in}|}
  \hline
& Euler pole & Euler pole & rotation & note and \\
Terrane & longitude & latitude & angle & citation \\
\hline
Greenland & -118.5 & 67.5 & -13.8 & Cenozoic separation of Greenland from Laurentia associated with opening of Baffin Bay and the Labrador Sea \citep{Roest1989a} \\
\hline
Scotland & 161.9 & 78.6 & -31.0 & Reconstruction of Atlantic Ocean opening following \cite{Torsvik2017a} \\
\hline
Svalbard & 305.0 & 81.0 & -68 & Rotate Svalbard to Laurentia in fit that works well with East Greenland basin according to \cite{Maloof2006a}\\
\hline
\end{tabular}
}
\label{tab:terrane_rotations}
\end{table}

Through the Proterozoic, there are intervals where there are abundant paleomagnetic poles that constrain Laurentia's position and intervals when the record is sparse (shown colored by age in Fig. \ref{fig:Laurentia_poles}). To further visualize the temporal coverage of the poles and to summarize the motion, implied paleolatitudes for an interior point on Laurentia are shown in Figure \ref{fig:Laurentia_paleolatitude}. The ages of the utilized paleomagnetic poles are also shown in comparison to the simplified summary of tectonic events in Figure \ref{fig:tectonic_history}. Both collisional and extensional tectonism can result in the formation of lithologies that can be used to develop paleomagnetic poles either as a result of basin formation, magmatism or both. In addition, intraplate magmatism resulting from plume-related large-igneous provinces (LIPs) can lead to paleomagnetic poles in periods that are otherwise characterized by tectonic quiescence (e.g. the ca. 1267 Ma Mackenzie LIP; Fig. \ref{fig:tectonic_history}). Intracontinental rifts have led to the highest density of poles both in the case of the ca. 1.4 Ga Belt-Purcell Supergroup and the ca. 1.1 Ga Midcontinent Rift (Figs. \ref{fig:Laurentia_map} and \ref{fig:tectonic_history}). The quality and resolution of the record from the Midcontinent Rift is aided by the voluminous magmatism that occurred in conjunction with basin formation that enables the development of a well-calibrated apparent polar wander path \citep{Swanson-Hysell2019a}. The late Tonian Period also has a number of poles including the Gunbarrel LIP (ca. 780 Ma) and Franklin LIP (ca. 720 Ma), as well as similarly-aged sedimentary rocks from western Laurentia basins \citep{Eyster2020a}. Overall, there is internal consistency among the paleomagnetic poles within intervals for which there is high-resolution coverage. These data result in progressive paths (Fig. \ref{fig:Laurentia_poles}) such as ascending to the ca. 1140 to 1110 Ma apex of the Logan Loop \citep{Robertson1971a}, down the ca. 1110 to 1080 Ma Keweenawan Track \citep{Swanson-Hysell2019a} to the ca. 980 Ma Grenville Loop \citep{McWilliams1975a} prior to a temporal gap before the late Tonian (ca. 775 to 720 Ma) path \citep{Eyster2020a}.

@@

An exception to this overall agreement between coeval poles from Greenland and mainland Laurentia occurs ca. 1382 Ma. There are poles of this age from Greenland associated with the Zig-Zag Dal basalts and related intrusions \citep{Marcussen1983a, Abrahamsen1987a}. However, these poles are in a distinct location from poles of similar age associated with the Belt-Purcell Supergroup (e.g. the McNamara Formation and Pilcher/Garnet Range and Libby Formations; \citealp{Elston2002a}). Additionally, the older Belt-Purcell Supergroup poles form a more southerly population than time-equivalent poles from elsewhere in Laurentia such as the Mistastin Pluton. There are potential complications associated with the Belt-Purcell Supergroup being exposed within thrust sheets with significant Mesozoic and Cenozoic deformation. However, vertical axis rotations of the Belt region are not able to bring the Belt poles into agreement with those from Laurentia or Greenland nor is translation away from the craton. Another potential complication is that the remanence used for the development of the Belt-Purcell Supergroup poles resides in hematite. As a result, there is the potential for inclination-flatting within the sedimentary rocks from which poles are developed. However, applying a moderate inclination factor of $f=0.6$ also does not bring the poles into congruence with the Zig-Zag Dal basalts. There is the potential that the hematite could be the result of post-depositional oxidation --- the remanence of the Purcell lavas pole is also held by hematite such that it is a chemical remanent magnetization (potentially acquired soon after eruption) rather than being a thermal remanent magnetization held by magnetite \citep{Elston2002a}. However the overall coherency of the pole directions from the Belt-Purcell Supergroup and the presence of reversals as interpreted from antipodal directions has been taken as evidence that the remanence is primary \citep{Elston2002a}. At present, it is unclear which poles are a better representation of Laurentia's position ca. 1400 Ma.

The interval of the record when there are the most significant inconsistencies between poles of similar age is the Ediacaran Period at the end of the Neoproterozoic Era (Figs. \ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitude}).  Between 583 and 565 Ma, paleomagnetic poles imply both low-latitude and high-latitude positions of Laurentia  (Fig. \ref{fig:Laurentia_paleolatitude}). This conflicting record is a longstanding problem and has led to the presentation of both high-latitude and low-latitude Laurentia paleogeographic reconstructions at the time (e.g. \citealp{Pisarevsky2001a,Li2008a}). One explanation for these variable pole positions is that they are the result of large-scale oscillatory true polar wander in the Ediacaran where rapid rotation of the entire silicate Earth influenced poles in Baltica and West Africa as well \citep{McCausland2007a, Robert2017a}. Paleodirectional data from single feldspar crystals from the Sept-\^Iles layered intrusion led \cite{Bono2015a} to interpret the lower inclination (and therefore lower latitude) direction from the intrusion (the one included as the ca. 565 Ma Sept-\^Iles pole in Table 2; \citealp{Tanczyk1987a}) as the primary thermal remanent magnetization.  \cite{Bono2015a} interpreted steeper directions also recovered from the intrusives as the result of remagnetization. They suggested that other steep magnetizations from Ediacaran Laurentia plutonic rocks, such as that observed in the ca. 583 Ma Baie des Mountons complex (the A group of \cite{McCausland2011a} in Table 2), are also the result of remagnetization. The lower inclination Baie des Mountons complex B Group directions result in a pole that is indistinguishable from the lower inclination  Sept-\^Iles intrusives pole. Another possibility discussed in the literature is that the lack of congruency between poles in this time interval is due to a particularly weak and non-dipolar geomagnetic field \citep{Abrajevitch2010a, Halls2015a, Bono2019a}. Data from the ca. 585 Ma Grenville dyke swarm of Laurentia, that are interpreted as primary, reveal $\sim$90\textdegree\ differences in direction within dikes dated within 2.5 $\pm$ 0.9 million years of one another \citep{Halls2015a}. The rates of $>$26\textdegree /Myr ($>$288 cm/yr) implied if these data are interpreted as resulting from plate motion or true polar wander were considered as dynamically implausible by \cite{Halls2015a} leading the authors to favor a deviation from axial dipolar behavior as the explanation for disparate Ediacaran directions. Estimates of magnetic paleointensity in these Grenville dikes are anomalously weak \citep{Thallner2020a} as are data from coeval volcanics in Ukraine \citep{Shcherbakova2019a} which could support an anomalous deviation from stable axial dipolar geomagnetic field behavior at the time as interpreted by \cite{Halls2015a}.  Regardless of mechanism, the Ediacaran data stand out as anomalous relative to the coherency of the rest of the poles in the compiled record for Laurentia (Fig. \ref{fig:Laurentia_paleolatitude}).

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../Figures/Fig6_Laurentia_reconstructions.pdf}
\caption{\textbf{Paleogeographic reconstructions of Laurentia at time intervals through the Proterozoic.} These reconstructions use the simple Laurentia pole interpolation model that is shown in Figure \ref{fig:Laurentia_paleolatitude} to reconstruct the tectonic elements of \cite{Whitmeyer2007a} shown in Figure \ref{fig:Laurentia_map}. Modern coastlines are maintained in these polygons so that the rotated orientations can be interpreted by the reader in comparison to Figure \ref{fig:Laurentia_map}. Paleomagnetic poles within 25 million years of each reconstruction time are plotted. All reconstructions have poles within such a time frame that provide constraints with the exception of the 850 Ma reconstruction which is shown faintly given this relative uncertainty in Laurentia's position. The colors follow Figure \ref{fig:Laurentia_map} where the light grey represents Archean provinces, dark grey represents Paleoproterozoic collisional orogens, and pink/orange represents Paleoproterozoic/Mesoproterozoic accretionary orogens and granitoid intrusions.}
\label{fig:Laurentia_reconstructions}
\end{figure*}

@@

There is an overall lack of models in the literature for the Proterozoic with published continuous rotation parameters that can be compared to the compilation of paleomagnetic poles presented herein. The approach in the community for many years has been to publish models as snapshots at time intervals presented in figures without publishing continuous rotation parameters, although some studies have published the Euler rotations associated with specified times. With the further adoption of software tools such as GPlates, there has been significant progress in the publication of continuous paleogeographic models constrained by paleomagnetic poles through the Phanerozoic (540 Ma to present; e.g. \citealp{Torsvik2017a}).

An exception to the paucity of published continuous paleogeographic models for the Precambrian is the Neoproterozoic model of \cite{Merdith2017b} which is shown in comparison to the constraints for Laurentia in Figure \ref{fig:Laurentia_paleolatitude}. The extent to which the implied position of Laurentia in \cite{Merdith2017b} is consistent with the compiled paleomagnetic constraints can be visualized in Figure \ref{fig:Laurentia_paleolatitude}. As noted above, the development of such models is challenging and researchers need to simultaneously balance many constraints. The focus here is on the extent to which this model satisfies the available paleomagnetic poles for Laurentia. The model does not honor the Grenville loop (e.g. Laurentia going to moderately high southerly latitudes ca. 1000 Ma), which is a striking departure from the paleomagnetic record and standard paleogeographic models. Additionally, the implemented plate motion of Laurentia in the \cite{Merdith2017b} model strays from the younger poles of the Keweenawan Track and does not honor the Franklin LIP pole \citep{Denyszyn2009b} despite its `A' Nordic rating (Fig. \ref{fig:Laurentia_paleolatitude}). The Franklin pole is taken to be a key constraint at the Tonian/Cryogenian boundary that provides evidence both for the supercontinent Rodinia being equatorial and for ice sheets associated with the Sturtian glaciation having extended to equatorial latitudes \citep{Macdonald2010a}.

There are more published models that show snapshots and publish rotation parameters associated with particular time intervals, such as the Rodinia model of \cite{Li2008a} and the Mesoproterozoic model of \cite{Pisarevsky2014b}, without providing parameters for a continuous model. The position for Laurentia implied by the Euler poles given for the model snapshots of these studies are shown in Figure \ref{fig:Laurentia_paleolatitude} and can be compared to the compiled record. The figure also shows the continuous implied position of Laurentia from the late Mesoproterozoic into the early Paleozoic from the model of \citet{Li2013a} (although the model parameters were not published with that study they have now been made available by the authors). This paleogeographic model implements large oscillations including ones in the Ediacaran that result from an interpretation that steep inclinations are the result of rapid motion of Laurentia from low to high latitudes and back again. The rates of Laurentia's motion associated with these models are also summarized in Figure \ref{fig:Laurentia_paleolatitude}. Over much of the record, Laurentia's pole positions can be satisfied through motion of the continent at rates of $<$10 cm/yr with intervals of more rapid motion such as during the Keweenawan Track and in the Paleozoic (Fig. \ref{fig:Laurentia_paleolatitude}).
